Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-09-21 MS/ms.Rmd
Added survey statement to abstract and conclusion

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We formalise effect size measures (e.g., the temperature coefficient, $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances opening new possibilities. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental designs, and environments. Importantly, sampling variances for many effect sizes commonly used by comparative physiologists (e.g., CTmax, $Q_{10}$ etc) can be formalized and powerful meta-analytic models can be easily applied to these effect measures (Figure \@ref(fig:fig0)). While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, factors such as temperature and dosage differences across studies may generate obvious sources of heterogeneity that may not be of prime interest.  Instead, comparative physiologists might simply want to focus on biological drivers of effect size variability and want an effect size or analytic approach that allows them to remove these ‘nuisances’. Here, we provide a set of tools to deal with ‘nuisance heterogeneity’ (Figure \@ref(fig:fig1)) by reinventing standard effect sizes and/or using mean centered nuisance variables in meta-regression models. Estimating complex meta-regression models with a multitude of fixed effects might result, however, in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by using an appropriate effect size and associated sampling variance while other biologically interesting moderators are fit in a meta-regression model might be desired. Indeed, this may even be a necessity to simplify the model. The set of tools we describe provide clarification around importance and limitations of heterogeneity. We also hope to provide new ways to help comparative physiologists communicate effect measures more richly to guide our understanding and decisions around pressing global challenges.

>>>>

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We assess the prevelance of nuisance hetrogeneity in the comparative physiology literature -- showing it is common and often not accounted for in analyses. We then formalise effect size measures (e.g., the temperature coefficient, $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances opening new possibilities. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental designs, and environments. Importantly, sampling variances for many effect sizes commonly used by comparative physiologists (e.g., CTmax, $Q_{10}$ etc) can be formalized and powerful meta-analytic models can be easily applied to these effect measures (Figure \@ref(fig:fig0)). While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, factors such as temperature and dosage differences across studies may generate obvious sources of heterogeneity that may not be of prime interest. Instead, comparative physiologists might simply want to focus on biological drivers of effect size variability and want an effect size or analytic approach that allows them to remove these ‘nuisances’. Surveying the comparative physiology literature shows that nuisance heterogenity is common, and is often not completely dealt with in a meta-analysis. Here, we provide a set of tools to deal with ‘nuisance heterogeneity’ (Figure \@ref(fig:fig1)) by reinventing standard effect sizes and/or using mean centered nuisance variables in meta-regression models. Estimating complex meta-regression models with a multitude of fixed effects might result, however, in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by using an appropriate effect size and associated sampling variance while other biologically interesting moderators are fit in a meta-regression model might be desired. Indeed, this may even be a necessity to simplify the model. The set of tools we describe provide clarification around importance and limitations of heterogeneity. We also hope to provide new ways to help comparative physiologists communicate effect measures more richly to guide our understanding and decisions around pressing global challenges.
